Conversations in time: interactive visualization to explore structured temporal data

### 2021-03-01 reviews/response.Rmd

---
title: Response to reviewers
author: RJournal 2020-171
output:
  UOARmdTemplates::memo:
    includes:
      in_header: preamble2.tex
fontfamily: mathpazo
---

We thank the reviewers for their comments which have improved this paper.

I've incorporated all feedback from reviewers regarding the package, except for adding unit tests. I'd like to include unit tests for code quality, but as far as I know, it's tricky to enable unit tests for interactive graphics.

Regarding the article, the point by point description of changes are below: the reviewers' comments are in red and our response is in black.

@@

# Reviewer 1 comments {-}

> Please mention that tidyverts also contains the tsibbletalk package

The core tidyverts only contains {tsibble}, {fable}, and {feasts}. The {tsibbletalk} packages is NOT part of the tidyverts suite.

@@

Thanks. We corrected "user-supplied session id" to "user-supplied id string".

# Reviewer 2 comments {-}

> Within the article, I would like to see more discussion of the actual linking implementation. I believe the package is using the R6 class and creating a subclass of SharedData from 'crosstalk', but it would be nice to see this explained within the article itself. A little note on why this is the preferred solution would also be of interest to readers

The actual linking is done through 'crosstalk'. We don't implement the linking rather than subclassing of its R6 object. I've added a note on why we prefer 'crosstalk' (or R6) as "This is an R6 object driving data transmission across multiple views, due to its mutable and lightweight properties." in the first paragraph of the section "Using a shared temporal data object for interactivity".

### 2021-07-18 tsibbletalk.Rmd
done with revision
minor tweak for publication
tweaks for publication

---
title: "Conversations in time: interactive visualisation to explore structured temporal data"
author:
- name: Earo Wang
  affiliation: The University of Auckland
  address: Department of Statistics
  email: earo.wang@auckland.ac.nz
- name: Dianne Cook
  affiliation: Monash University
  address: Department of Econometrics and Business Statistics
  email: dicook@monash.edu
bibliography: tsibbletalk.bib
abstract: >
  Temporal data often has a hierarchical structure, defined by categorical variables describing different levels, such as political regions or sales products. Nesting of categorical variables produces a hierarchical structure. The \CRANpkg{tsibbletalk} package is developed to allow a user to interactively explore temporal data, relative to the nested or crossed structures. It can help to discover differences between category levels, and uncover interesting periodic or aperiodic slices. The package implements a shared `tsibble` object that allows for linked brushing between coordinated views, and a shiny module that aids in wrapping time lines for seasonal patterns. The tools are demonstrated using two data examples: domestic tourism in Australia and pedestrian traffic in Melbourne.
preamble: |
  \usepackage{subfig}
output: rticles::rjournal_article
---

@@

Temporal data typically arrives as a set of many observational units measured over time. Some variables may be categorical, containing a hierarchy in the collection process, that may be measurements taken in different geographic regions, or types of products sold by the one company. Exploring these multiple features can be daunting. Ensemble graphics [@unwin_ensemble_2018] bundle multiple views of a data set together into one composite figure. These provide an effective approach for exploring and digesting many different aspects of temporal data. Adding interactivity to the ensemble can greatly enhance the exploration process.

@@

There is a long history of interactive data visualisation research and corresponding systems. Within R, the systems can be roughly divided into systems utilising web technology and those that don't.

@@

Systems without the web technology include \CRANpkg{grDevices}, \CRANpkg{loon} [@R-loon], based on Tcl/Tk, and \pkg{cranvas} [@xie_reactive_2014] based on Qt. They offer a wide array of pre-defined interactions, such as selecting and zooming, to manipulate plots via mouse action, keyboard strokes, and menus. The \pkg{cranvastime} package [@cheng_enabling_2016] is an add-on to \pkg{cranvas}, which provides specialised interactions for temporal data, such as wrapping and mirroring.

The techniques implemented in work described in this paper utilise web technology, including \CRANpkg{crosstalk}, \CRANpkg{plotly}, and R \CRANpkg{shiny}.

@@

The \CRANpkg{tsibbletalk} package introduces a shared tsibble instance built on a `tsibble`. This allows for seamless communication between different plots of temporal data. The `as_shared_tsibble()` function turns a `tsibble` into a shared instance,  `SharedTsibbleData`, which is a subclass of `SharedData` from \CRANpkg{crosstalk}. This is the object driving data transmission across multiple views. The \CRANpkg{tsibbletalk} package aims to streamline interactive exploration of temporal data, with the focus of temporal elements and structured linking.

@@

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to categorical variable linking where selecting one or more observations in one category will broadcast to all other observations in this category. That is, linking is by key variables: within the time series plot, click on any data point, and  whole line will be highlighted in response. 
The `as_shared_tsibble()` uses `tsibble`'s `key` variables to achieve these types of linking.

@@

To illustrate nesting and crossing we use the `tourism_monthly` dataset [@tourism] packaged in \CRANpkg{tsibbletalk}. It contains monthly domestic overnight trips across Australia. The `key` is comprised of three identifying variables: `State`, `Region`, and `Purpose` (of trip), in particular `State` nesting of `Region`, together crossed with `Purpose`. This specification can be translated as follows:

@@

\noindent There is a three-level hierarchy: the root node is implicitly Australia, and geographically disaggregated to states and lower-level tourism regions. A new handy function `plotly_key_tree()` has been implemented to help explore the hierarchy. It interprets hierarchies in the shared tsibble's `spec` as a tree view, built with \CRANpkg{plotly}. The following code line produces the linked tree diagram (left panel of Figure \ref{fig:tourism-linking-fig}). The visual for the tree hierarchy detangles a group of related series and provides a bird's eye view of the data organisation.

@@

The tree plot provides the graphics skeleton, upon which the rest of data plots can be attached. In this example, small multiples of line plots are placed at the top right of Figure \ref{fig:tourism-linking-fig} to explore the temporal trend across regions, by trip purpose. The shared tsibble data can be directly piped into \CRANpkg{ggplot2} code to create this.

@@

These line plots are heavily overplotted. To tease apart structure in the multiple time series, the `features()` function computes interesting characteristics, including the measures of trend and seasonality. These are displayed in the scatterplot at bottom right, where one dot represents one series.

@@

There is one final step, to compose the three plots into an ensemble of coordinated views for exploration, shown in Figure \ref{fig:tourism-linking-fig}.  (This is the interactive realisation of Figure \ref{fig:highlight-retail}). <!-- Routine functions bring about new interaction with temporal data on the client side.-->

@@

1. A branch of the tree corresponding to Western Australia was first selected. (The names of the regions are a little odd, which is a quirk of the data set, but all four areas, Australia's South West, ...., correspond to tourist destinations in Western Australia. Hovering over the node on the branch brings up the state name.) This generated the response in the line plots and the scatterplot that coloured corresponding time series and points as blue. 
2. To enable persistent selection, in oder to compare regions or states, "Shift" and click on the tree was done, after switching the colour to red. This generated the response that points and time series corresponding to Sydney were highlighted in red. 
3. Hovering over the points brings up the label for Sydney.

The domestic tourism sees Sydney as one of the most popular destinations in realm of business and friends visiting over the years. Despite the relatively weaker performance in Western Australia, Australia's North West region sees a strongest upward trend in Business, bypassing Sydney in some years.

In summary, shared tsibble data nicely bridges between the \CRANpkg{crosstalk} and **tidyverts** ecosystems for temporal data using the common "key". The `as_shared_tsibble()` provides a symbolic user interface for effortless construction of a hybrid of hierarchical and categorical linking between plots. The `plotly_key_tree()` function, in turn, decodes the hierarchical specification to plot a tree for data overview and navigation, when accompanied by more detailed plots.

@@

An important aspect of temporal data is the time context. Time has cyclical structure, that may correspond to seasonal patterns to be discovered. The `index` component of the (shared) tsibble data forms the basis for exploring seasonality. To investigate for periodic or aperiodic patterns, series should be wrapped on themselves, where the index is broken into temporal components like quarter or day. We shall explore this with pedestrian traffic in Melbourne, Australia.

@@

The city of Melbourne has sensors installed at various locations, to record hourly counts of pedestrians, in order to capture daily rhythms of the downtown [@ped]. Figure \ref{fig:wrap-ped} shows the first five months of 2020 foot traffic at four different locations, for three different time slices, daily, weekly and full five months. Plot \ref{fig:wrap-ped-1} shows hourly counts from January to May on an absolute time line, facetted by locations. The stage 3 COVID-19 lockdown, on March 16, is marked by a change of colour. (The pre-lockdown period is coloured with dark green and lockdown with orange.) We can see a significant decline in foot traffic  at all four locations. QV Market is less affected probably because this is a major produce market, an essential service that continued to operate. Bourke St, a major shopping center, sees a gradual uptick in the last weeks of the period indicating that people were getting back into the shops.

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show slicing and wrapping of the series into daily and weekly sections, respectively. Multiple seasonalities pop out. There tends to be a daily pattern, especially visible at the main train station, Southern Cross Station. There is also a weekday vs weekend pattern, also most visible at Southern Cross Station. These seasonal patterns are still present during the lockdown, but the magnitude is greatly reduced. Numbers are also down at the produce market and the shopping center. Birrarung Marr is the most affected. This is the location of special events, and it is clear that these have completely disappeared during lockdown.

@@

This shiny module, decoupled to `tsibbleWrapUI()` and `tsibbleWrapServer()`, presents a clean interface and forms a reusable component that could be embedded in any shiny application. In general, a shiny module provides a vehicle for modularising shiny applications, relevant for both users and developers. As with all shiny modules, the first argument in both functions in \CRANpkg{tsibbletalk} requires a user-supplied id string that must be unique. The UI function `tsibbleWrapUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is the server function `tsibbleWrapServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, so that the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. For date-times (represented by `POSIXt`), the granularity ranges from fine "day" to a much coarser "year". The following code snippet generates Figure \ref{fig:wrap-ped}.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the slider incremented by 1-day units. The "play" button near the end of slider can automatically animate the slicing and dicing process, walking the viewer through all 24 hours of the 152 days. Alternatively, users can drag the slider to examine selected slices.

@@

The new shiny module exploits the temporal aspect for a tsibble object, available through the `index` attribute. It allows users to slide through relative periods to digest seasonal behaviours, with nimble user experience.

@@

The `as_shared_tsibble()` turns a `tsibble` object to a shared data class, with an option to express any nesting and crossing structures from the `key` attribute. If nesting is found in the data, the `plotly_key_tree()` creates an interactive hierarchical tree to help with data overview. This sets the stage for hierarchical and categorical linking between multiple views from one shared tsibble.

>>>>

---
title: "Conversations in time: interactive visualization to explore structured temporal data"
author:
- name: Earo Wang
  affiliation: The University of Auckland
  address: Department of Statistics
  email: earo.wang@auckland.ac.nz
- name: Dianne Cook
  affiliation: Monash University
  address: Department of Econometrics and Business Statistics
  email: dicook@monash.edu
bibliography: tsibbletalk.bib
abstract: >
  Temporal data often has a hierarchical structure, defined by categorical variables describing different levels, such as political regions or sales products. The nesting of categorical variables produces a hierarchical structure. The \CRANpkg{tsibbletalk} package is developed to allow a user to interactively explore temporal data, relative to the nested or crossed structures. It can help to discover differences between category levels, and uncover interesting periodic or aperiodic slices. The package implements a shared `tsibble` object that allows for linked brushing between coordinated views, and a shiny module that aids in wrapping timelines for seasonal patterns. The tools are demonstrated using two data examples: domestic tourism in Australia and pedestrian traffic in Melbourne.
preamble: |
  \usepackage{subfig}
output: rticles::rjournal_article
---

@@

Temporal data typically arrives as a set of many observational units measured over time. Some variables may be categorical, containing a hierarchy in the collection process, that may be measurements taken in different geographic regions, or types of products sold by one company. Exploring these multiple features can be daunting. Ensemble graphics [@unwin_ensemble_2018] bundle multiple views of a data set together into one composite figure. These provide an effective approach for exploring and digesting many different aspects of temporal data. Adding interactivity to the ensemble can greatly enhance the exploration process.

@@

There is a long history of interactive data visualization research and corresponding systems. Within R, the systems can be roughly divided into systems utilizing web technology and those that do not.

@@

Systems without the web technology include \CRANpkg{grDevices}, \CRANpkg{loon} [@R-loon], based on Tcl/Tk, and \pkg{cranvas} [@xie_reactive_2014] based on Qt. They offer a wide array of pre-defined interactions, such as selecting and zooming, to manipulate plots via mouse action, keyboard strokes, and menus. The \pkg{cranvastime} package [@cheng_enabling_2016] is an add-on to \pkg{cranvas}, which provides specialized interactions for temporal data, such as wrapping and mirroring.

The techniques implemented in the work described in this paper utilize web technology, including \CRANpkg{crosstalk}, \CRANpkg{plotly}, and R \CRANpkg{shiny}.

@@

The \CRANpkg{tsibbletalk} package introduces a shared tsibble instance built on a `tsibble`. This allows for seamless communication between different plots of temporal data. The `as_shared_tsibble()` function turns a `tsibble` into a shared instance,  `SharedTsibbleData`, which is a subclass of `SharedData` from \CRANpkg{crosstalk}. This is an R6 object driving data transmission across multiple views, due to its mutable and lightweight properties. The \CRANpkg{tsibbletalk} package aims to streamline interactive exploration of temporal data, with the focus of temporal elements and structured linking.

@@

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to categorical variable linking, where selecting one or more observations in one category will broadcast to all other observations in this category. That is, linking is by key variables: within the time series plot, click on any data point, and the whole line will be highlighted in response. The `as_shared_tsibble()` uses `tsibble`'s `key` variables to achieve these types of linking.

@@

To illustrate nesting and crossing we use the `tourism_monthly` dataset [@tourism] packaged in \CRANpkg{tsibbletalk}. It contains monthly domestic overnight trips across Australia. The `key` is comprised of three identifying variables: `State`, `Region`, and `Purpose` (of the trip), in particular `State` nesting of `Region`, crossed together with `Purpose`. This specification can be translated as follows:

@@

\noindent There is a three-level hierarchy: the root node is implicitly Australia, geographically disaggregated to states, and lower-level tourism regions. A new handy function `plotly_key_tree()` has been implemented to help explore the hierarchy. It interprets hierarchies in the shared tsibble's `spec` as a tree view, built with \CRANpkg{plotly}. The following code line produces the linked tree diagram (left panel of Figure \ref{fig:tourism-linking-fig}). The visual for the tree hierarchy detangles a group of related series and provides a bird's eye view of the data organization.

@@

The tree plot provides the graphics skeleton, upon which the rest of the data plots can be attached. In this example, small multiples of line plots are placed at the top right of Figure \ref{fig:tourism-linking-fig} to explore the temporal trend across regions by the trip purpose. The shared tsibble data can be directly piped into \CRANpkg{ggplot2} code to create this.

@@

These line plots are heavily overplotted. To tease apart structure in the multiple time series, the `features()` function computes interesting characteristics, including the measures of trend and seasonality. These are displayed in the scatterplot at the bottom right, where one dot represents one series.

@@

There is one final step, to compose the three plots into an ensemble of coordinated views for exploration, shown in Figure \ref{fig:tourism-linking-fig}.  (This is the interactive realization of Figure \ref{fig:highlight-retail}). <!-- Routine functions bring about new interaction with temporal data on the client side.-->

@@

1. A branch of the tree corresponding to Western Australia was first selected. (The names of the regions are a little odd, which is a quirk of the data set, but all four areas, Australia's South West, ...., correspond to tourist destinations in Western Australia. Hovering over the node on the branch brings up the state name.) This generated the response in the line plots and the scatterplot that colored corresponding time series and points as blue. 
2. To enable persistent selection, in oder to compare regions or states, "Shift" and click on the tree was done, after switching the color to red. This generated the response that points and time series corresponding to Sydney were highlighted in red. 
3. Hovering over the points brings up the label for Sydney.

Domestic tourism sees Sydney as one of the most popular destinations in the realm of business and friends visiting over the years. Despite the relatively weaker performance in Western Australia, Australia's North West region sees a strongest upward trend in business, bypassing Sydney in some years.

In summary, shared tsibble data nicely bridges between the \CRANpkg{crosstalk} and **tidyverts** ecosystems for temporal data using the common "key". The `as_shared_tsibble()` provides a symbolic user interface for the effortless construction of a hybrid of hierarchical and categorical linking between plots. The `plotly_key_tree()` function, in turn, decodes the hierarchical specification to plot a tree for data overview and navigation, when accompanied by more detailed plots.

@@

An important aspect of temporal data is the time context. Time has a cyclical structure, that may correspond to seasonal patterns to be discovered. The `index` component of the (shared) tsibble data forms the basis for exploring seasonality. To investigate for periodic or aperiodic patterns, series should be wrapped on themselves, where the index is broken into temporal components like quarter or day. We shall explore this with pedestrian traffic in Melbourne, Australia.

@@

The city of Melbourne has sensors installed at various locations, to record hourly counts of pedestrians, in order to capture the daily rhythms of the downtown [@ped]. Figure \ref{fig:wrap-ped} shows the first five months of 2020 foot traffic at four different locations, for three different time slices, daily, weekly and full five months. Plot \ref{fig:wrap-ped-1} shows hourly counts from January to May on an absolute timeline, facetted by locations. The stage 3 COVID-19 lockdown, on March 16, is marked by a change of color. (The pre-lockdown period is colored with dark green and lockdown with orange.) We can see a significant decline in foot traffic  at all four locations. QV Market is less affected probably because this is a major produce market, an essential service that continued to operate. Bourke St, a major shopping center, sees a gradual uptick in the last weeks of the period indicating that people were getting back into the shops.

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show the slicing and wrapping of the series into daily and weekly sections, respectively. Multiple seasonalities pop out. There tends to be a daily pattern, especially visible at the main train station, Southern Cross Station. There is also a weekday vs weekend pattern, also most visible at Southern Cross Station. These seasonal patterns are still present during the lockdown, but the magnitude is greatly reduced. Numbers are also down at the produce market and the shopping center. Birrarung Marr is the most affected. This is the location of special events, and it is clear that these have completely disappeared during the lockdown.

@@

This shiny module, decoupled to `tsibbleWrapUI()` and `tsibbleWrapServer()`, presents a clean interface and forms a reusable component that could be embedded in any shiny application. In general, a shiny module provides a vehicle for modularising shiny applications, relevant for both users and developers. As with all shiny modules, the first argument in both functions in \CRANpkg{tsibbletalk} requires a user-supplied id string that must be unique. The UI function `tsibbleWrapUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is the server function `tsibbleWrapServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, so that the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. For date-times (represented by `POSIXt`), the granularity ranges from fine "day" to a much coarser "year". The following code snippet generates Figure \ref{fig:wrap-ped}. The creation of the `pedestrian20` data is available in supplementary R files.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the slider incremented by 1-day units. The "play" button near the end of the slider can automatically animate the slicing and dicing process, walking the viewer through all 24 hours of the 152 days. Alternatively, users can drag the slider to examine selected slices.

@@

The new shiny module exploits the temporal aspect for a tsibble object, available through the `index` attribute. It allows users to slide through relative periods to digest seasonal behaviors, with a nimble user experience.

@@

The `as_shared_tsibble()` turns a `tsibble` object to a shared data class, with an option to express any nesting and crossing structures from the `key` attribute. If nesting is found in the data, the `plotly_key_tree()` creates an interactive hierarchical tree to help with the data overview. This sets the stage for hierarchical and categorical linking between multiple views from one shared tsibble.
